---
title: "Compare within arm"
output: html_document
date: "2025-09-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
library(dplyr)
library(ggplot2)
```

```{r load-data, include=FALSE}
source("R/load_and_process_data.R")
dir.create("results/figures", recursive = TRUE, showWarnings = FALSE)
```

```{r colors-labels}
arm_colors <- c("arm_A" = "#4E79A7", "arm_B" = "#F28E2B")
arm_labels <- c("arm_A" = "Combination treatment arm", "arm_B" = "Monotherapy arm")

diff_variant_lables <- c("BL" = "Only baseline", "both" = "Maintained", "PD" = "Only acquired")

focal_copy_number_colors <- c(
    "acquired_gain" = "#D08159",
    "stable_gain"   = "#F2A97F",
    "acquired_loss" = "#8E6BAF",
    "stable_loss"   = "#C7AEDB"
)

timepoint_arm_A_colors      <- c("BL" = "#A6C8E0", "PD" = "#4E79A7")
timepoint_arm_B_colors      <- c("BL" = "#F8C99B", "PD" = "#F28E2B")
timepoint_labels            <- c("BL" = "Baseline", "PD" = "Progression")
timepoint_both_arms_colors  <- c("arm_A_BL" = "#A6C8E0", "arm_A_PD" = "#4E79A7",
                                 "arm_B_BL" = "#F8C99B", "arm_B_PD" = "#F28E2B")
timepoint_both_arms_labels  <- c("arm_A_BL" = "Baseline (arm A)", "arm_A_PD" = "Progression (arm A)",
                                 "arm_B_BL" = "Baseline (arm B)", "arm_B_PD" = "Progression (arm B)")
```

## Somatic mutation burden

```{r mutation-burden}
somatic_mutation_burden_BL <- somatic_tso500_ALL_trace_variants_paired |>
    filter(diff_variant == "BL") |>
    group_by(patnr, arm) |>
    summarize(n_somatic_mutations = n(), .groups = "drop")

wilcox.test(n_somatic_mutations ~ arm, data = somatic_mutation_burden_BL)

somatic_mutation_burden_BL_both <- somatic_tso500_ALL_trace_variants_paired |>
    filter(diff_variant %in% c("BL", "both")) |>
    group_by(patnr, arm) |>
    summarize(n_somatic_mutations = n(), .groups = "drop")

wilcox.test(n_somatic_mutations ~ arm, data = somatic_mutation_burden_BL_both)

somatic_mutation_burden_PD <- somatic_tso500_ALL_trace_variants_paired |>
    filter(diff_variant == "PD") |>
    group_by(patnr, arm) |>
    summarize(n_somatic_mutations = n(), .groups = "drop")

wilcox.test(n_somatic_mutations ~ arm, data = somatic_mutation_burden_PD)
```

## McNemar tests — BL vs PD within each arm

```{r mcnemar-function}
mcnemar_gene <- function(x) {
    tbl <- matrix(
        c(x$none, x$BL, x$PD, x$both),
        nrow = 2, byrow = TRUE
    )
    test <- mcnemar.test(tbl)
    tibble::tibble(
        p_value    = test$p.value,
        chi_squared = test$statistic,
        arm  = x$arm,
        none = x$none,
        BL   = x$BL,
        PD   = x$PD,
        both = x$both
    )
}
```

```{r contingency-within-arm}
contingency_within_arm_variant_counts <- somatic_tso500_ALL_trace_variants_paired |>
    select(patnr, arm, genename, diff_variant) |>
    distinct() |>
    group_by(genename, patnr) |>
    mutate(n = n()) |>
    ungroup() |>
    mutate(diff_variant = ifelse(n == 1, diff_variant, "both")) |>
    select(patnr, arm, genename, diff_variant) |>
    distinct() |>
    select(!patnr) |>
    group_by(arm, genename, diff_variant) |>
    summarize(n = n(), .groups = "drop") |>
    tidyr::pivot_wider(names_from = c(diff_variant), values_from = n, values_fill = 0) |>
    mutate(
        none = case_when(
            arm == "arm_A" ~ 21 - (both + BL + PD),
            arm == "arm_B" ~ 14 - (both + BL + PD)
        )
    )
```

```{r mcnemar-tests}
arm_A_BL_vs_PD <- contingency_within_arm_variant_counts |>
    filter(arm == "arm_A") |>
    group_by(genename) |>
    group_modify(~mcnemar_gene(.x)) |>
    ungroup() |>
    arrange(p_value)

arm_B_BL_vs_PD <- contingency_within_arm_variant_counts |>
    filter(arm == "arm_B") |>
    group_by(genename) |>
    group_modify(~mcnemar_gene(.x)) |>
    ungroup() |>
    arrange(p_value)
```

```{r gene-ordering-oncoprint}
gene_ordering_oncoprint <- contingency_within_arm_variant_counts |>
    tidyr::pivot_wider(names_from = arm,
                       values_from = c(both, BL, PD, none), values_fill = 0) |>
    mutate(
        BL           = both_arm_A + both_arm_B + BL_arm_A + BL_arm_B,
        BL_perc      = BL / {{ n_patients_cohort }},
        PD           = both_arm_A + both_arm_B + PD_arm_A + PD_arm_B,
        PD_perc      = PD / {{ n_patients_cohort }},
        PD_armA      = both_arm_A + PD_arm_A,
        PD_armA_perc = PD_armA / 21,
        PD_armB      = both_arm_B + PD_arm_B,
        PD_armB_perc = PD_armB / 14
    ) |>
    arrange(desc(PD_armA_perc)) |>
    filter(PD_armA_perc > 0.1) |>
    mutate(gene_ordering = row_number()) |>
    select(genename, gene_ordering)

arm_A_BL_vs_PD |>
    right_join(gene_ordering_oncoprint, by = join_by(genename)) |>
    mutate(p_adj = p.adjust(p_value, method = "fdr"))

arm_B_BL_vs_PD |>
    right_join(gene_ordering_oncoprint, by = join_by(genename)) |>
    mutate(p_adj = p.adjust(p_value, method = "fdr"))
```

## Oncoprints — BL vs PD mutation frequency per arm

```{r oncoprint-arm-a}
arm_A_BL_vs_PD |>
    right_join(gene_ordering_oncoprint, by = join_by(genename)) |>
    mutate(
        total = none + BL + PD + both,
        PD    = (PD + both) / total,
        BL    = (BL + both) / total
    ) |>
    tidyr::pivot_longer(cols = c(PD, BL), names_to = "timepoint", values_to = "percentage") |>
    ggplot(aes(x = reorder(genename, gene_ordering), y = percentage, fill = timepoint)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.5)) +
    coord_cartesian(ylim = c(0, 1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_fill_manual(values = timepoint_arm_A_colors) +
    labs(
        title = "Baseline and Progression mutations in combination treatment arm",
        x = "Gene", y = "Percentage of patients with mutation", fill = "Timepoint"
    ) +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'bottom',
        legend.direction = 'horizontal',
        text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )
ggsave("results/figures/compare_within_arm_oncoprint_arm_a.pdf", plot = last_plot(), width = 9, height = 5)

```

```{r oncoprint-arm-b}
arm_B_BL_vs_PD |>
    right_join(gene_ordering_oncoprint, by = join_by(genename)) |>
    mutate(
        total = none + BL + PD + both,
        PD    = (PD + both) / total,
        BL    = (BL + both) / total
    ) |>
    tidyr::pivot_longer(cols = c(PD, BL), names_to = "timepoint", values_to = "percentage") |>
    ggplot(aes(x = reorder(genename, gene_ordering), y = percentage, fill = timepoint)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.5)) +
    coord_cartesian(ylim = c(0, 1)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_fill_manual(values = timepoint_arm_B_colors) +
    labs(
        title = "Baseline and Progression mutations in Monotherapy arm",
        x = "Gene", y = "Percentage of patients with mutation", fill = "Timepoint"
    ) +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'bottom',
        legend.direction = 'horizontal',
        text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )
ggsave("results/figures/compare_within_arm_oncoprint_arm_b.pdf", plot = last_plot(), width = 9, height = 5)

```

## Focal copy number — McNemar tests within arm

```{r cn-mcnemar}
contingency_copy_number_genes <- copy_number_genes |>
    right_join(gene_ordering_focal_cna, by = join_by(genename)) |>
    filter(alt == "deletion") |>
    select(patnr, arm, genename, diff_variant_focal_cna) |>
    distinct() |>
    select(!patnr) |>
    group_by(arm, genename, diff_variant_focal_cna) |>
    summarize(n = n(), .groups = "drop") |>
    tidyr::pivot_wider(names_from = c(diff_variant_focal_cna), values_from = n, values_fill = 0) |>
    mutate(
        none = case_when(
            arm == "arm_A" ~ 21 - (both + BL + PD),
            arm == "arm_B" ~ 14 - (both + BL + PD)
        )
    )

contingency_copy_number_genes |>
    filter(arm == "arm_A") |>
    group_by(genename) |>
    group_modify(~mcnemar_gene(.x))

contingency_copy_number_genes |>
    filter(arm == "arm_B") |>
    group_by(genename) |>
    group_modify(~mcnemar_gene(.x))
```

## Focal CNA bar charts per arm

```{r cn-barchart-arm-a}
copy_number_genes |>
    right_join(gene_ordering_focal_cna, by = join_by(genename)) |>
    group_by(arm, timepoint, genename, gene_ordering, alt) |>
    summarize(n = n(), .groups = "drop") |>
    mutate(percentage = ifelse(arm == "arm_A", n / 21, n / 14)) |>
    filter(arm == "arm_A") |>
    select(!n) |>
    tidyr::pivot_wider(names_from = timepoint, values_from = percentage, values_fill = 0) |>
    tidyr::pivot_longer(cols = c(PD, BL), names_to = "timepoint", values_to = "percentage") |>
    ggplot(aes(x = reorder(genename, gene_ordering, decreasing = TRUE),
               y = percentage, fill = timepoint)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.5)) +
    facet_grid("alt") +
    coord_cartesian(ylim = c(0, 1)) +
    scale_fill_manual(values = timepoint_arm_A_colors) +
    labs(
        title = "Oncoprint focal CNA's arm A, BL vs PD",
        x = "Gene", y = "Percentage of patients with focal CNA", fill = "Timepoint"
    ) +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'bottom',
        legend.direction = 'horizontal',
        text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )
ggsave("results/figures/compare_within_arm_cn_barchart_arm_a.pdf", plot = last_plot(), width = 9, height = 5)

```

```{r cn-barchart-arm-b}
copy_number_genes |>
    right_join(gene_ordering_focal_cna, by = join_by(genename)) |>
    group_by(arm, timepoint, genename, gene_ordering, alt) |>
    summarize(n = n(), .groups = "drop") |>
    mutate(percentage = ifelse(arm == "arm_A", n / 21, n / 14)) |>
    filter(arm == "arm_B") |>
    select(!n) |>
    tidyr::pivot_wider(names_from = timepoint, values_from = percentage, values_fill = 0) |>
    tidyr::pivot_longer(cols = c(PD, BL), names_to = "timepoint", values_to = "percentage") |>
    ggplot(aes(x = reorder(genename, gene_ordering, decreasing = TRUE),
               y = percentage, fill = timepoint)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.5)) +
    facet_grid("alt") +
    coord_cartesian(ylim = c(0, 1)) +
    scale_fill_manual(values = timepoint_arm_B_colors) +
    labs(
        title = "Oncoprint focal CNA's arm B, BL vs PD",
        x = "Gene", y = "Percentage of patients with focal CNA", fill = "Timepoint"
    ) +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'bottom',
        legend.direction = 'horizontal',
        text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )
ggsave("results/figures/compare_within_arm_cn_barchart_arm_b.pdf", plot = last_plot(), width = 9, height = 5)

```

## Focal CNA burden

```{r cn-burden}
focal_copy_number_burden <- copy_number_genes_wide |>
    filter(!cn_category %in% c("gain_normalized", "loss_normalized")) |>
    group_by(patnr, arm) |>
    summarize(n_focal_cn_per_patient = n(), .groups = "drop")

wilcox.test(n_focal_cn_per_patient ~ arm, data = focal_copy_number_burden)
```

Total focal CNAs per patient (stacked by category):
```{r cn-total-per-patient}
copy_number_genes_wide |>
    filter(!cn_category %in% c("gain_normalized", "loss_normalized")) |>
    group_by(patnr, arm, cn_category) |>
    summarize(n = n(), .groups = "drop") |>
    mutate(n = ifelse(is.na(n), 0, n)) |>
    right_join(patient_order_somatic_variants, by = join_by(patnr, arm)) |>
    ggplot(aes(x = reorder(patnr, patient_ordering, decreasing = TRUE), y = n, fill = cn_category)) +
    geom_bar(stat = "identity") +
    facet_grid(~arm, scales = "free", space = "free", labeller = labeller(arm = arm_labels)) +
    scale_fill_manual(
        values   = focal_copy_number_colors,
        na.value = "grey80",
        breaks   = c("acquired_gain", "stable_gain", "acquired_loss", "stable_loss", NA)
    ) +
    labs(title = "Total focal CNA's and types", y = "Count", x = "Patient ID", fill = "Alterations") +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'bottom',
        legend.direction = 'horizontal',
        text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )
ggsave("results/figures/compare_within_arm_cn_total_per_patient.pdf", plot = last_plot(), width = 10, height = 5)
```

## Somatic alteration frequency per patient

```{r alteration-frequency-helpers}
# shared helper: builds "both" category without consequence filter,
# keeping one row per variant (PD timepoint) with CCF > 3% at both timepoints.
# the original script never applied protein_coding_consequences to "both" variants.
build_somatic_both <- function(data) {
    data |>
        filter(diff_variant == "both", status == "Somatic") |>
        group_by(patnr, positionComplete) |>
        mutate(n = n()) |>
        filter(timepoint == "PD" | n != 2) |>
        ungroup() |>
        select(patnr, sampleName_tso500, timepoint, arm, genename, diff_variant, status,
               positionComplete, position, proteinchange, varianttype, refcall, altcall,
               consequence, vaf_BL, vaf_PD, cn_BL, cn_PD, ccf_BL, ccf_PD, depth,
               tumorFractionEstimate) |>
        filter(ccf_BL > 0.03 & ccf_PD > 0.03)
}

# shared columns selected for BL/PD variants
variant_cols <- c("patnr", "sampleName_tso500", "timepoint", "arm", "genename",
                  "diff_variant", "status", "positionComplete", "position",
                  "proteinchange", "varianttype", "refcall", "altcall", "consequence",
                  "vaf_BL", "vaf_PD", "cn_BL", "cn_PD", "ccf_BL", "ccf_PD", "depth",
                  "tumorFractionEstimate")

freq_plot <- function(data, title_str) {
    plot_data <- data |>
        dplyr::group_by(patnr, arm, diff_variant, varianttype) |>
        dplyr::summarise(n = dplyr::n(), .groups = "drop")

    totals <- plot_data |>
        dplyr::group_by(patnr, arm, diff_variant) |>
        dplyr::summarise(total_n = sum(n), .groups = "drop")

    ggplot(plot_data,
           aes(x = reorder(patnr, n, decreasing = TRUE), y = n, fill = varianttype)) +
        geom_bar(stat = "identity") +
        geom_text(
            data = totals,
            aes(x = reorder(patnr, total_n, decreasing = TRUE), y = total_n, label = total_n),
            inherit.aes = FALSE, vjust = -0.4, size = 4
        ) +
        facet_grid(
            diff_variant ~ arm,
            scales   = "free",
            space    = "free",
            labeller = labeller(arm = arm_labels, diff_variant = diff_variant_lables)
        ) +
        coord_cartesian(ylim = c(1, 200)) +
        labs(title = title_str, x = "Patient ID", y = "n alterations", fill = "") +
        theme_light() +
        ggplot2::theme(
            legend.position  = 'bottom',
            legend.direction = 'horizontal',
            text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
            axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
            axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
            strip.background = element_rect(fill = "lightgray"),
            strip.text       = element_text(color = "black", size = 10),
            axis.title       = element_text(size = 15, face = "bold"),
            plot.title       = element_text(size = 15, face = "bold")
        )

}
```

### All variants

```{r alteration-frequency-all}
somatic_bl_fig1 <- trace_variants_paired |>
    filter(diff_variant == "BL", status == "Somatic") |>
    select(all_of(variant_cols))

somatic_pd_fig1 <- trace_variants_paired |>
    filter(diff_variant == "PD", status == "Somatic") |>
    select(all_of(variant_cols))

variants_fig1 <- bind_rows(somatic_bl_fig1, somatic_pd_fig1, build_somatic_both(trace_variants_paired)) |>
    arrange(patnr, genename, positionComplete) |>
    mutate(mutation_id = paste0(genename, "_", positionComplete))

freq_plot(variants_fig1,
          "Frequency of somatic alterations, per patient")
ggsave("results/figures/compare_within_arm_alteration_frequency_all.pdf", plot = last_plot(), width = 10, height = 8)
```

### All protein-altering variants

```{r alteration-frequency-protein-altering}
somatic_bl_fig2 <- trace_variants_paired |>
    filter(diff_variant == "BL", status == "Somatic",
           consequence %in% protein_coding_consequences) |>
    select(all_of(variant_cols))

somatic_pd_fig2 <- trace_variants_paired |>
    filter(diff_variant == "PD", status == "Somatic",
           consequence %in% protein_coding_consequences) |>
    select(all_of(variant_cols))

variants_fig2 <- bind_rows(somatic_bl_fig2, somatic_pd_fig2, build_somatic_both(trace_variants_paired)) |>
    arrange(patnr, genename, positionComplete) |>
    mutate(mutation_id = paste0(genename, "_", positionComplete))

freq_plot(variants_fig2,
          "Frequency of somatic protein altering alterations, per patient")
ggsave("results/figures/compare_within_arm_alteration_frequency_protein_altering.pdf", plot = last_plot(), width = 10, height = 8)
```

### All protein-altering variants, CCF > 3%

```{r alteration-frequency-ccf-cutoff}
somatic_bl_fig3 <- trace_variants_paired |>
    filter(diff_variant == "BL", status == "Somatic",
           consequence %in% protein_coding_consequences, ccf_BL > 0.03) |>
    select(all_of(variant_cols))

somatic_pd_fig3 <- trace_variants_paired |>
    filter(diff_variant == "PD", status == "Somatic",
           consequence %in% protein_coding_consequences, ccf_PD > 0.03) |>
    select(all_of(variant_cols))

variants_fig3 <- bind_rows(somatic_bl_fig3, somatic_pd_fig3, build_somatic_both(trace_variants_paired)) |>
    arrange(patnr, genename, positionComplete) |>
    mutate(mutation_id = paste0(genename, "_", positionComplete))

freq_plot(variants_fig3,
          "Frequency of somatic protein altering alterations (CCF>3%), per patient")
ggsave("results/figures/compare_within_arm_alteration_frequency_ccf_cutoff.pdf", plot = last_plot(), width = 10, height = 8)
```

## Variant type summary

```{r variant-type-summary}
variants_fig3 |>
    dplyr::group_by(varianttype) |>
    dplyr::summarize(n = dplyr::n(), .groups = "drop")
```

## Mutation burden — baseline vs progression

```{r mutation-burden-wilcox}
n_baseline_mutations <- variants_fig3 |>
    filter(diff_variant %in% c("BL", "both")) |>
    group_by(patnr, arm) |>
    summarize(n = n(), .groups = "drop")

wilcox.test(n ~ arm, data = n_baseline_mutations, exact = FALSE)

n_baseline_mutations |>
    group_by(arm) |>
    summarize(median_n = median(n), .groups = "drop")

n_progression_mutations <- variants_fig3 |>
    filter(diff_variant %in% c("PD", "both")) |>
    group_by(patnr, arm) |>
    summarize(n = n(), .groups = "drop")

wilcox.test(n ~ arm, data = n_progression_mutations, exact = FALSE)

n_progression_mutations |>
    group_by(arm) |>
    summarize(median_n = median(n), .groups = "drop")
```

```{r}

```

