---
title: "Compare arms"
output: html_document
date: "2025-09-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
library(dplyr)
library(ggplot2)
```

```{r load-data, include=FALSE}
source("R/load_and_process_data.R")
dir.create("results/figures", recursive = TRUE, showWarnings = FALSE)
```

```{r colors-labels}
arm_colors <- c("arm_A" = "#4E79A7", "arm_B" = "#F28E2B")
arm_labels  <- c("arm_A" = "Combination (AI + CDK4/6i)", "arm_B" = "Monotherapy (AI)")

mutation_label <- c(
    "acquired"             = "Aquired",
    "acquired_polyclonal"  = "Aquired multiple",
    "acquired_maintained"  = "Aquired & maintained",
    "maintained"           = "Maintained",
    "maintained_polyclonal" = "Maintained multiple"
)

mutation_label_colors <- c(
    "acquired"             = "#D16A6B",
    "acquired_polyclonal"  = "#A05555",
    "acquired_maintained"  = "#F5CDD1",
    "maintained"           = "#3A7350",
    "maintained_polyclonal" = "#B0BA93"
)

focal_copy_number_labels <- c(
    "acquired_gain" = "Acquired gain",
    "stable_gain"   = "Stable gain",
    "acquired_loss" = "Aquired loss",
    "stable_loss"   = "Stable loss"
)

focal_copy_number_colors <- c(
    "acquired_gain" = "#D08159",
    "stable_gain"   = "#F2A97F",
    "acquired_loss" = "#8E6BAF",
    "stable_loss"   = "#C7AEDB"
)
```

## Somatic mutation burden

Test mutation burden between arms at progression:
```{r mutation-burden}
somatic_mutation_burden <- somatic_tso500_PD_trace_variants_paired |>
    group_by(patnr, arm) |>
    summarize(n_somatic_mutations = n(), .groups = "drop")

wilcox.test(n_somatic_mutations ~ arm, data = somatic_mutation_burden)
```

Per-gene Fisher tests between arms (PD variants):
```{r per-gene-tests}
per_gene_tests |>
    mutate(total_mut = n_mut_A + n_mut_B) |>
    filter(total_mut > 4) |>
    pull(genename)

per_gene_tests |> arrange(desc(n_mut_B))
```

Wilcoxon tests per top gene:
```{r top-gene-wilcoxon}
n_top_mut_compare <- bind_rows(lapply(top_genes, function(top_gene) {
    meta_pd <- meta_data |> filter(timepoint == "PD")
    mutation_count_patients <- somatic_tso500_PD_trace_variants_paired |>
        filter(genename == {{ top_gene }}) |>
        group_by(patnr, arm) |>
        summarize(n = n(), .groups = "drop") |>
        right_join(meta_pd, by = join_by(patnr, arm)) |>
        mutate(n = ifelse(is.na(n), 0, n))
    mutation_count_patients_armA <- mutation_count_patients |> filter(arm == "arm_A")
    mutation_count_patients_armB <- mutation_count_patients |> filter(arm == "arm_B")
    test_res <- wilcox.test(mutation_count_patients_armA$n, mutation_count_patients_armB$n)
    tibble(
        genename       = top_gene,
        p_value        = test_res$p.value,
        median_mut_armA = mean(mutation_count_patients_armA$n),
        median_mut_armB = mean(mutation_count_patients_armB$n)
    )
}))

n_top_mut_compare |> mutate(p_adj = p.adjust(p_value, method = "fdr"))
```

Fisher test for RB1 (hardcoded from manual curation):
```{r rb1-fisher}
mat <- matrix(c(7, 0, 14, 21), nrow = 2,
              dimnames = list(Arm = c("Combination", "Monotherapy"),
                              RB1 = c("Event", "No event")))
fisher.test(mat)
```

## Oncoplot — tornado chart

```{r oncoplot-tornado}
n_mutations_per_patient <- somatic_tso500_PD_trace_variants_paired |>
    right_join(top_genes_alteration, by = join_by(genename)) |>
    left_join(patient_order_somatic_variants, by = join_by(patnr, arm)) |>
    left_join(mutation_labels, by = join_by(patnr, genename, arm)) |>
    group_by(patnr, arm, genename, gene_order, total_mut, mutation_label) |>
    summarize(n = n(), .groups = "drop") |>
    mutate(perc = n / total_mut)

percentage_labels <- n_mutations_per_patient |>
    group_by(arm, genename, gene_order) |>
    summarize(n_mutations_per_arm = length(unique(patnr)), .groups = "drop") |>
    mutate(
        perc_mutations_per_arm = round(ifelse(arm == "arm_A", n_mutations_per_arm / 21, -n_mutations_per_arm / 14) * 100),
        perc_label = paste0(abs(perc_mutations_per_arm), "%"),
        x = ifelse(arm == "arm_A", perc_mutations_per_arm + 11, perc_mutations_per_arm - 11)
    )

n_mutations_per_patient |>
    group_by(arm, genename, gene_order, mutation_label) |>
    summarize(n_mutations_per_arm = length(unique(patnr)), .groups = "drop") |>
    mutate(perc = round(ifelse(arm == "arm_A", n_mutations_per_arm / 21, -n_mutations_per_arm / 14) * 100)) |>
    ggplot() +
    geom_bar(aes(y = reorder(genename, gene_order, decreasing = TRUE), x = perc, fill = mutation_label),
             stat = "identity") +
    coord_cartesian(xlim = c(-100, 100)) +
    scale_x_continuous(breaks = c(-100, -50, 0, 50, 100),
                       labels = c("100%", "50%", "0%", "50%", "100%")) +
    scale_fill_manual(labels = mutation_label, values = mutation_label_colors) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.3) +
    geom_text(data = percentage_labels,
              aes(x = x, y = reorder(genename, gene_order, decreasing = FALSE), label = perc_label)) +
    labs(title = "", x = "% Patients", y = "Genes", fill = "Alterations") +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'none',
        text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )
ggsave("results/figures/compare_arms_oncoplot_tornado.pdf", plot = last_plot(), width = 7, height = 6)
```

## Oncoplot — tile plot per patient

```{r oncoplot-tiles}
n_mutations_per_patient_at_PD <- somatic_tso500_PD_trace_variants_paired |>
    right_join(top_genes_alteration, by = join_by(genename)) |>
    left_join(patient_order_somatic_variants, by = join_by(patnr, arm)) |>
    left_join(mutation_labels, by = join_by(patnr, genename, arm)) |>
    group_by(patnr, arm, genename, diff_variant) |>
    summarize(n_mututation_at_PD = n(), .groups = "drop") |>
    mutate(n_mututation_at_PD = ifelse(diff_variant == "both", ceiling(n_mututation_at_PD / 2), n_mututation_at_PD)) |>
    group_by(patnr, arm, genename) |>
    summarize(n_mututation_at_PD = sum(n_mututation_at_PD), .groups = "drop")

somatic_tso500_PD_trace_variants_paired |>
    right_join(top_genes_alteration, by = join_by(genename)) |>
    left_join(mutation_labels, by = join_by(patnr, genename, arm)) |>
    bind_rows(missing_patients) |>
    tidyr::complete(tidyr::nesting(patnr, arm), tidyr::nesting(genename, gene_order)) |>
    filter(!is.na(genename)) |>
    left_join(patient_order_somatic_variants, by = join_by(patnr, arm)) |>
    left_join(n_mutations_per_patient_at_PD, by = join_by(patnr, arm, genename)) |>
    mutate(arm = factor(arm, levels = c("arm_B", "arm_A"))) |>
    ggplot(aes(
        x = reorder(patnr, patient_ordering, decreasing = TRUE),
        y = reorder(genename, gene_order, decreasing = TRUE))
    ) +
    ggplot2::geom_tile(aes(fill = as.factor(mutation_label)), color = "gray", linewidth = 0.1) +
    geom_text(aes(label = round(n_mututation_at_PD, 2)), color = "white", size = 2.5, na.rm = TRUE) +
    scale_fill_manual(labels = mutation_label, values = mutation_label_colors,
                      na.value = "white", na.translate = TRUE) +
    facet_grid(~arm, scales = "free", space = "free", labeller = labeller(arm = arm_labels)) +
    labs(title = "Maintained and acquired mutations", x = "Patient ID", y = "Genes", fill = "Alterations") +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'bottom',
        legend.direction = 'horizontal',
        text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )
ggsave("results/figures/compare_arms_oncoplot_tiles.pdf", plot = last_plot(), width = 11, height = 6)
```

## Focal copy number

Correlation between focal (TSO500) and chromosomal (ichorCNA) copy number:
```{r cn-correlation}
copy_number_genes |>
    ggplot(aes(x = cn_tumor_gene_int, y = ichorCNA_cn)) +
    geom_jitter(width = 0.5, height = 0.5, alpha = 0.5) +
    geom_vline(xintercept = 1) +
    geom_hline(yintercept = 1) +
    geom_abline(slope = 1, intercept = 0) +
    ggpubr::stat_cor(p.accuracy = 0.001, method = "spearman", size = 4, cor.coef.name = "rho") +
    labs(
        title = "Focal and chromosomal copy number of copy number altered genes",
        y = "Focal copy number",
        x = "Chromosomal copy number"
    ) +
    theme_light() +
    theme(
        plot.title       = element_text(size = 15, face = "bold"),
        axis.title       = element_text(size = 15, face = "bold"),
        axis.text.x      = element_text(size = 10),
        legend.position  = "bottom",
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10)
    )
ggsave("results/figures/compare_arms_copy_number_correlation.pdf", plot = last_plot(), width = 6, height = 5)

```

Per-gene Fisher tests for focal copy number alterations between arms:
```{r cn-fisher-tests}
per_gene_copy_number_test <- copy_number_genes_wide |>
    filter(diff_variant_focal_cna != "BL") |>
    select(patnr, arm, genename) |>
    distinct() |>
    select(!patnr) |>
    group_by(arm, genename) |>
    summarize(n = n(), .groups = "drop") |>
    tidyr::pivot_wider(names_from = arm, values_from = n, values_fill = 0) |>
    group_by(genename) |>
    group_modify(~fisher_gene(.x)) |>
    ungroup() |>
    arrange(p_value)

top_genes_copy_number <- per_gene_copy_number_test |>
    mutate(total_cn_alteration = n_mut_A + n_mut_B) |>
    right_join(gene_ordering_focal_cna, by = join_by(genename)) |>
    arrange(gene_ordering, genename) |>
    mutate(p_adj = p.adjust(p_value, method = "BH"))

top_genes_copy_number |>
    arrange(p_value) |>
    mutate(p_adj = p.adjust(p_value, method = "BH")) |>
    arrange(gene_ordering, genename)
```

Focal CNA oncoplot — tile per patient:
```{r cn-oncoplot}
patients_with_mut_in_top_focal_cn <- copy_number_genes_wide |>
    filter(diff_variant_focal_cna != "BL") |>
    right_join(top_genes_copy_number, by = join_by(genename)) |>
    pull(patnr) |>
    unique()

missing_patients_focal_cn <- meta_data |>
    filter(!patnr %in% patients_with_mut_in_top_focal_cn, timepoint == "PD")

# Patient ordering based on CN alteration burden (descending total events)
cn_patient_ordering <- copy_number_genes_wide |>
    filter(diff_variant_focal_cna != "BL") |>
    right_join(top_genes_copy_number |> select(genename, gene_ordering), by = join_by(genename)) |>
    select(patnr, arm, genename) |>
    distinct() |>
    mutate(has_cn = 1L) |>
    tidyr::pivot_wider(names_from = genename, values_from = has_cn, values_fill = 0L) |>
    right_join(meta_data |> filter(timepoint == "PD") |> select(patnr, arm), by = join_by(patnr, arm)) |>
    mutate(across(any_of(top_genes_copy_number$genename), ~ifelse(is.na(.), 0L, .))) |>
    mutate(n_cn_total = rowSums(across(any_of(top_genes_copy_number$genename)), na.rm = TRUE)) |>
    arrange(desc(n_cn_total), across(rev(any_of(top_genes_copy_number$genename)), ~ desc(.x))) |>
    mutate(cn_patient_ordering = row_number()) |>
    select(patnr, arm, cn_patient_ordering)

# Full grid: all PD patients × all top CN genes — ensures patients with no
# events appear as empty columns rather than being silently dropped
full_cn_grid <- tidyr::crossing(
    meta_data |> filter(timepoint == "PD") |> select(patnr, arm),
    top_genes_copy_number |> select(genename, gene_ordering)
)

focal_cn_oncoplot <- copy_number_genes_wide |>
    filter(diff_variant_focal_cna != "BL") |>
    right_join(top_genes_copy_number, by = join_by(genename)) |>
    right_join(full_cn_grid, by = join_by(patnr, arm, genename, gene_ordering)) |>
    left_join(cn_patient_ordering, by = join_by(patnr, arm)) |>
    mutate(arm = factor(arm, levels = c("arm_B", "arm_A"))) |>
    ggplot2::ggplot(aes(
        x = reorder(patnr, cn_patient_ordering, decreasing = FALSE),
        y = reorder(genename, gene_ordering, decreasing = FALSE))
    ) +
    ggplot2::geom_tile(aes(fill = as.factor(cn_category)), na.rm = TRUE, color = "black", linewidth = 0.1) +
    geom_text(aes(label = round(cn_PD, 2)), color = "white", size = 2.5, na.rm = TRUE) +
    scale_fill_manual(values = focal_copy_number_colors, na.value = "white",
                      labels = focal_copy_number_labels, na.translate = TRUE) +
    facet_grid(~arm, scales = "free", space = "free", labeller = labeller(arm = arm_labels)) +
    labs(title = "Retained and acquired focal copy number alterations",
         x = "Patient ID", y = "Genes", fill = "Alterations") +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'bottom',
        legend.direction = 'horizontal',
        text             = ggplot2::element_text(size = 9, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    ) +
    guides(fill = guide_legend(nrow = 1, byrow = FALSE))

focal_cn_oncoplot
ggsave("results/figures/compare_arms_focal_cn_oncoplot.pdf", plot = focal_cn_oncoplot, width = 11, height = 6)

```

Tornado chart of focal CNA frequencies between arms:
```{r cn-tornado}
percentage_labels_focal_cn <- copy_number_genes_wide |>
    filter(diff_variant_focal_cna != "BL") |>
    right_join(top_genes_copy_number, by = join_by(genename)) |>
    right_join(patient_order_somatic_variants, by = join_by(patnr, arm)) |>
    select(patnr, arm, genename, cn_category, total_cn_alteration, gene_ordering) |>
    distinct() |>
    group_by(genename, arm, total_cn_alteration, cn_category, gene_ordering) |>
    summarize(n = n(), .groups = "drop") |>
    mutate(perc = ifelse(arm == "arm_A", n / 21, -n / 14)) |>
    group_by(genename, arm, gene_ordering) |>
    summarize(perc_total = round(sum(perc) * 100), .groups = "drop") |>
    mutate(
        perc_label = paste0(abs(perc_total), "%"),
        x = ifelse(arm == "arm_A", perc_total / 100 + 0.16, perc_total / 100 - 0.16)
    ) |>
    filter(!is.na(genename))

focal_cn_frequency <- copy_number_genes_wide |>
    filter(diff_variant_focal_cna != "BL") |>
    right_join(top_genes_copy_number, by = join_by(genename)) |>
    right_join(patient_order_somatic_variants, by = join_by(patnr, arm)) |>
    select(patnr, arm, genename, cn_category, total_cn_alteration, gene_ordering) |>
    distinct() |>
    group_by(genename, arm, total_cn_alteration, cn_category, gene_ordering) |>
    summarize(n = n(), .groups = "drop") |>
    mutate(perc = ifelse(arm == "arm_A", n / 21, -n / 14)) |>
    filter(!is.na(genename)) |>
    ggplot() +
    geom_bar(aes(
        y    = reorder(genename, gene_ordering, decreasing = FALSE),
        x    = perc,
        fill = factor(cn_category, levels = c("acquired_gain", "stable_gain", "acquired_loss", "stable_loss"))
    ), stat = "identity") +
    coord_cartesian(xlim = c(-1, 1)) +
    scale_x_continuous(breaks = c(-1, -0.5, 0, 0.5, 1),
                       labels = c("100%", "50%", "0%", "50%", "100%")) +
    scale_fill_manual(values = focal_copy_number_colors, na.value = "grey80",
                      labels = focal_copy_number_labels) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.3) +
    geom_text(data = percentage_labels_focal_cn,
              aes(x = x, y = reorder(genename, gene_ordering, decreasing = FALSE), label = perc_label), size = 3) +
    labs(title = "Frequency of focal copy number alterations",
         x = "% Patients", y = "Genes", fill = "Alterations") +
    theme_light() +
    ggplot2::theme(
        legend.position  = 'bottom',
        legend.direction = 'horizontal',
        text             = ggplot2::element_text(size = 8, family = 'Helvetica', face = 'bold'),
        axis.text.x      = element_text(hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 8),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    ) +
    guides(fill = guide_legend(nrow = 1, byrow = FALSE))

focal_cn_frequency
ggsave("results/figures/compare_arms_focal_cn_tornado.pdf", plot = focal_cn_frequency, width = 7, height = 6)
```


## Combined oncoplot: tiles + tornado

```{r oncoplot-combined, fig.width=9, fig.height=5}
oncoplot_tile <- somatic_tso500_PD_trace_variants_paired |>
    right_join(top_genes_alteration, by = join_by(genename)) |>
    left_join(mutation_labels, by = join_by(patnr, genename, arm)) |>
    bind_rows(missing_patients) |>
    tidyr::complete(tidyr::nesting(patnr, arm), tidyr::nesting(genename, gene_order)) |>
    filter(!is.na(genename)) |>
    left_join(patient_order_somatic_variants, by = join_by(patnr, arm)) |>
    left_join(n_mutations_per_patient_at_PD, by = join_by(patnr, arm, genename)) |>
    mutate(arm = factor(arm, levels = c("arm_B", "arm_A"))) |>
    ggplot(aes(
        x = reorder(patnr, patient_ordering, decreasing = TRUE),
        y = reorder(genename, gene_order, decreasing = TRUE))
    ) +
    ggplot2::geom_tile(aes(fill = as.factor(mutation_label)), color = "gray", linewidth = 0.1) +
    geom_text(aes(label = round(n_mututation_at_PD, 2)), color = "white", size = 2.5, na.rm = TRUE) +
    scale_fill_manual(labels = mutation_label, values = mutation_label_colors,
                      na.value = "white", na.translate = TRUE) +
    facet_grid(~arm, scales = "free", space = "free", labeller = labeller(arm = arm_labels)) +
    labs(title = "Maintained and acquired mutations", x = "Patient ID", y = "Genes", fill = "Alterations") +
    theme_light() +
    ggplot2::theme(
        legend.position  = "bottom",
        legend.direction = "horizontal",
        text             = ggplot2::element_text(size = 9, family = "Helvetica", face = "bold"),
        axis.text.x      = element_text(angle = 45, hjust = 1, size = 9),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )

frequency_bar <- n_mutations_per_patient |>
    group_by(arm, genename, gene_order, mutation_label) |>
    summarize(n_mutations_per_arm = length(unique(patnr)), .groups = "drop") |>
    mutate(perc = round(ifelse(arm == "arm_A", n_mutations_per_arm / 21, -n_mutations_per_arm / 14) * 100)) |>
    ggplot() +
    geom_bar(aes(y = reorder(genename, gene_order, decreasing = TRUE), x = perc, fill = mutation_label),
             stat = "identity") +
    coord_cartesian(xlim = c(-100, 100)) +
    scale_x_continuous(breaks = c(-100, -50, 0, 50, 100),
                       labels = c("100%", "50%", "0%", "50%", "100%")) +
    scale_fill_manual(labels = mutation_label, values = mutation_label_colors) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.3) +
    geom_text(data = percentage_labels,
              aes(x = x, y = reorder(genename, gene_order, decreasing = FALSE), label = perc_label),
              size = 2) +
    labs(x = "% Patients", y = "", fill = "Alterations") +
    theme_light() +
    ggplot2::theme(
        legend.position  = "none",
        text             = ggplot2::element_text(size = 7, family = "Helvetica", face = "bold"),
        axis.text.x      = element_text(size = 9),
        axis.text.y      = element_blank(),
        axis.title.y     = element_blank(),
        axis.title.x     = ggtext::element_textbox_simple(width = NULL, halign = .5),
        strip.background = element_rect(fill = "lightgray"),
        strip.text       = element_text(color = "black", size = 10),
        axis.title       = element_text(size = 15, face = "bold"),
        plot.title       = element_text(size = 15, face = "bold")
    )

combined_oncoplot <- patchwork::wrap_plots(oncoplot_tile, frequency_bar, design = "AAAAAAAAAABBB#")
combined_oncoplot
ggsave("results/figures/compare_arms_oncoplot_combined.pdf", plot = combined_oncoplot, width = 9, height = 5)
```


## Combined focal CN oncoplot: tiles + tornado

```{r focal-cn-combined, fig.width=9, fig.height=5}
combined_focal_cn_oncoplot <- patchwork::wrap_plots(
    focal_cn_oncoplot,
    focal_cn_frequency +
        labs(title = "", y = "") +
        ggplot2::theme(
            legend.position = "none",
            axis.text.y     = element_blank(),
            axis.title.y    = element_blank()
        ),
    design = "AAAAAAAAAABBB#"
)
combined_focal_cn_oncoplot
ggsave("results/figures/compare_arms_focal_cn_combined.pdf",
       plot = combined_focal_cn_oncoplot, width = 9, height = 5)
```


```{r}

```

